---
title: "Predicting 2015 Citibike Trips"
author: "Gabrielle Martinez (Pace University)"
date: "6/19/2020"
output: html_document
---

```{r setup, include=FALSE}
library(here)
library(scales)
library(broom)
library(modelr)
library(tidyverse)
library(lubridate)

theme_set(theme_bw())


knitr::opts_chunk$set(echo = TRUE)
```

## Using a model trained on 2014 data to predict 2015 trends

First let's load the 2015 data for cleaning and reformatting. 

```{r load}

load('trips2015.RData')

head(trips_per_day2015)

```

Next let's load our model and a function to add holiday data to `trips_per_day2015`

```{r format}

load('predict_citibike_model.RData')

#add holiday and factor columns
modifiy_trips_data(trips_per_day2015) -> trips_per_day2015

head(trips_per_day2015)

ggplot(trips_per_day2015, aes(x=ymd, y=num_trips)) +
  geom_point() +
  geom_smooth()

```
`trips_perday2015` now has 5 more columns. `holiday` states the holiday for the day and `isholiday` states if the day is a holiday or not (1,0). `israining` states if it rained that day if the precipitation variable >= 0.1 inches. `day_of_week` tells what day of the week it is and based on that`isweekday` states if the day is a weekday or a weekend. 


##Predicting 2015 trips per day

Using the following model we try to predict the trips per day 

```{r}

tidy(model4)
glance(model4)

rmse(model4, trips_per_day2015)

trips_per_day2015 %>%
  add_predictions(model4) -> trips_per_day2015

ggplot(trips_per_day2015, aes(x=ymd, y=pred)) +
  geom_point()


```








